Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2023-02-21 Edith Zemanick/Sweat Test/sweat_test_analysis.Rmd
Update sweat_test_analysis.Rmd
flow chart finished

```{r setup, include=FALSE}
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Windows") {
  home_dir <- "C:/Users/timvigers/OneDrive - The University of Colorado Denver/Vigers/CF/Edith Zemanick/Sweat Test"
  source("C:/Users/timvigers/Documents/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Linux") {
  home_dir <- "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Darwin") {
  home_dir <- "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/genotype_class_severity.R")
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in%
      c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

@@

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` <- as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` <- as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(numeric.test = "kwt", numeric.stats = c("Nmiss", "median", "q1q3"))
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
```

@@

```{r results='asis'}
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) # First measure for everyone
t1 <- tableby(
  Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + fe(geno_class) +
    Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```

>>>>

```{r setup, include=FALSE}
library(DiagrammeR)
library(readxl)
library(arsenal)
library(tidyverse)
library(pROC)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Windows") {
  home_dir <- "C:/Users/timvigers/OneDrive - The University of Colorado Denver/Vigers/CF/Edith Zemanick/Sweat Test"
  source("C:/Users/timvigers/Documents/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Linux") {
  home_dir <- "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/get_genotype_class_severity.R")
} else if (Sys.info()["sysname"] == "Darwin") {
  home_dir <- "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Edith Zemanick/Sweat Test"
  source("~/GitHub/CF/genotype_class_severity.R")
}
knitr::opts_knit$set(root.dir = home_dir)
rm(home_dir)
```

```{r}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
n1 = nrow(df)
n2 = length(unique(df$PatientID))
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
df$Diagnosis <- diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)]
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
exclude2 = c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)
no_dem = unique(df$PatientID[df$Demographics.Available!="TRUE" | is.na(df$Demographics.Available) | df$Sex==""])
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in% exclude2),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

@@

# Difference between arms
df$`Cl Diff. Between Left and Right Arm` <- as.numeric(df$Cl_Left) - as.numeric(df$Cl_Right)
df$`Na Diff. Between Left and Right Arm` <- as.numeric(df$Na_Left) - as.numeric(df$Na_Right)
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
high_ratio = unique(which(df$`Na/Cl Ratio`>2))
high_cl = unique(which(df$Cl>160))
high_na = unique(which(df$Na>160))
df = df[-unique(c(high_ratio,high_cl,high_na)),]
# Sweat chloride group
df$`Sweat Chloride Group` = cut(df$Cl,c(-Inf,30,60,Inf),right = F,
                                labels = c("< 30","30-59",">60"))
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(numeric.test = "kwt", numeric.stats = c("Nmiss", "median", "q1q3"))
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
# First measure for everyone
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1) 
```

# Flowchart

```{r}
grViz("digraph {

# initiate graph
graph [layout = dot]

# global node settings
node [shape = rectangle]

# label nodes
data1 [label = '@@1']
data2 [label = '@@2']
ineligible [label = '@@3']
demo  [label = '@@4']
highna  [label = '@@5']
highcl  [label = '@@6']
highnacl  [label = '@@7']
review [label = '@@8']
analysis [label = '@@9']

# edge definitions with the node IDs
data1  -> data2
data2 -> ineligible
ineligible -> {demo,highna,highcl,highnacl,review}
data2  -> analysis
highnacl -> analysis [style=invis]
}
[1]: paste0('Sweat tests performed \\n ','n = ',n1)
[2]: paste0('Unique participants \\n ','n = ',n2)
[3]: paste0('Ineligible \\n ','n = ',n2-nrow(t1_df))
[4]: paste0('No demographics \\n ','n = ',length(no_dem))
[5]: paste0('Na > 160 mmol/L \\n ','n = ',length(high_na))
[6]: paste0('Cl > 160 mmol/L \\n ','n = ',length(high_cl))
[7]: paste0('Na:Cl > 2 \\n ','n = ',length(high_ratio))
[8]: paste0('Other \\n ','n = ',1)
[9]: paste0('Analysis cohort \\n ','n = ',nrow(t1_df))
")
```

@@

```{r results='asis'}
t1 <- tableby(
  Diagnosis ~ Sex + Age.at.Test.in.Years + fe(geno_risk) + fe(geno_class) +
    Amount + Cl + Na + K + `Na/Cl Ratio`,
  data = t1_df, control = mycontrols, simulate.p.value = T
)
summary(t1,
  pfootnote = T,
  labelTranslations = labels
)
```
